WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-17 articles/2014/Article-WSClean/wsclean.tex
Work on article

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{E-mail:
\url{andre.offringa@anu.edu.au}}, B.~McKinley$^{1,2}$, N.~Hurley-Walker, 
F.~H.~Briggs, \newauthor
J.~Rhee, L.~Feng, D.~Kaplan, A.~R.~Neben, J.~D.~Hughes, MWA commissioning team\newauthor
and GLEAM members where appropriate \& MWA builders list
\\
$^{1}$RSAA, Australian National University, Mt Stromlo Observatory, via Cotter Road, Weston, ACT 2611, Australia \\
$^{2}$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO) \\
}

@@

\begin{abstract}
We present a new imager implementation that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm. On MWA data, we find it to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution. The performance dependencies of CASA's $w$-projection and WSClean are analysed and analytical functions are derived that describe the required computing cost for both imagers. The difference in computing cost is mainly caused by the efficiency of our new imager on MWA's 27\degree field of view. We predict the computing costs for several other configurations and estimate that our imager is approximately a factor of 2--3 faster on other arrays. Especially imaging SKA observations will become challenging, for which we predict that our new algorithm is still at least 30 times slower than is budgetted. We find that the $w$-snapshot algorithm has only a minor benefit. The source code of our new imager is publicly released.
\end{abstract}

@@

Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with a two-dimensional Fast Fourier Transformation (FFT). Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}. This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes that by nature are wide-field instruments.

There are several methods to deal with the w-term during imaging: facetting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, w-projection \citep{wprojection-cornwell}, w-stacking \citep{widefield-imaging-ska-cornwell} and w-snapshots \citep{widefield-imaging-ska-cornwell}.

New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}). The Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution. However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's w-projection algorithm. Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.

Another option exists for imaging MWA data: the Real-Time System (RTS, \citealt{rts-mwa}). This has been designed as an efficient calibration and imaging pipeline specifically for MWA data. It can use GPUs for improving efficiency, and uses a w-projection kernel during imaging to correct for the w-term. The kernel can be kept small, because the data is processed in snapshots that have been phase-centered to zenith. The RTS can not perform deconvolution as of yet.

To reach high dynamic ranges, it can be necessary to deal with direction-dependent effects (DDEs). This is especially true for widefield telescopes. One way to handle DDEs is by using the a-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
One such DDE is the effect of the ionosphere. For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure \citep{lonsdale-calibration-approaches-ionosphere}. This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data. The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, w-projection and w-stacking \citep{awimager-2013}.

Once the Square-Kilometre Array (SKA) begins its operation, the required computational power for wide-field imaging will become an even bigger challenge. \citet{widefield-imaging-ska-cornwell} argues that the w-snapshots algorithm is the most efficient approach for the SKA.

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's w-projection implementation, and is able to succesfully image any MWA observation. To obtain the increase in speed, the implementation uses the w-stacking method for correcting the w-terms, optionally combined with w-snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''. Our new imaging implementation, which we believe to be the fastest generic widefield imager available, is publicly released\footnote{TODO add URL}.

@@

\section{The w-stacking technique} \label{sec:wstacking}
In this section, we will describe the $w$-stacking algorithm from a mathematical point of view.

@@

To derive the w-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \times \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \times \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \times \\ \notag
& e^{2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_0$ to $w_e$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_e - w_0\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \times \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~grid samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~apply the direction dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeat this for all $w$-values and add the results together; v)~apply the final scaling.

In practice, the final scaling will be different from $\left(w_e - w_0\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples. Therefore, the number of added samples and their weight have to be taken into account. Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction dependent effects. In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds. The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-values samples. In this case we can therefore calculate the image for $w<0$ from the image with $w>0$. That allows us to set $w_0$ to the minimum absolute $w$-value, which requires two times fewer layers. In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence the inverse FFT is always performed as a complex-to-complex transform.

@@

\subsection{Discretization of w} \label{sec:gridding-w}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/aliasing-example-annotated}
\caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field. WSClean was set to use 12 $w$-layers. The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers. Sources are 1~Jy (red circles), ghost sources are approximately 0.2~Jy. Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce more ghost sources. }
\label{fig:aliasing-example}
\end{center}
\end{figure}
While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of w-layers that need to be processed. For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian. This results in the constraint
\begin{equation} \label{eq:minimum-w-distance}
\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
\end{equation}
This suggest that a uniform discretization in $w$ is optimal (TODO this is in contrast to what Cornwell 2008 says -- Cornwell mailed me he made a mistake... Do we need to explicitly point that out here :-/ ?). From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
\begin{equation} \label{eq:nwlayers-bound}
 N_\textrm{wlayers} \gg 2\pi \left(w_e - w_0\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
\end{equation}
Actual values for the right-hand side can differ a lot. The value of $w_e - w_0$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image. For the MWA, a typical value of $w_e - w_0$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$. For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$. This implies that tens of $w$-layers are required at zenith and hundreds at lower elevations. The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in the next sections.

@@

\subsection{Time complexity of w-stacking} \label{sec:time-complexity}
It is useful to analyse the time complexity of w-stacking and compare it with w-projection, to understand which algorithm performs better in a given situation. We will use the following symbols: $N_\textrm{wlay}$ is the number of w-layers for w-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each dimension, $N_\textrm{vis}$ is the number of visibilities and $N_\textrm{kern}$ is the size of the antialiasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the w-kernel for w-projection, $w_e$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.

\begin{table}
 \caption{Scaling of the computational cost for various imaging steps} \label{tbl:computational-cost-per-operation}
 \begin{tabular}{lll}
   \textbf{Operation} & \textbf{W-stacking} & \textbf{W-projection} \\
   \hline\hline
   Fourier transform(s) & $N_\textrm{wlay} N^2_\textrm{pix} \log N_\textrm{pix}$ & $N^2_\textrm{pix} \log N_\textrm{pix}$ \\
   W-term corrections   & $N_\textrm{wlay} N^2_\textrm{pix}$ & $N_\textrm{vis} N^2_\textrm{wkern}$ \\
   Gridding & $N_\textrm{vis}N^2_\textrm{kern}$ & $N_\textrm{vis} N_\textrm{kern}^2$ \\
   \hline\hline
 \end{tabular}
\end{table}

Table~\ref{tbl:computational-cost-per-operation} shows how the computational costs scale for the operations that dominate the imaging in the w-stacking and w-projection algorithms. For comparison, we can assume the gridding kernel adds a constant term and the terms $N_\textrm{wlay}$ and $N_\textrm{wkern}$ follow approximately $w_e \sin \alpha_\textrm{FOV}$. The time complexity for w-stacking is then given by
\begin{equation}
\textrm{TC}_\textrm{wstacking}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} w_e \sin \alpha_\textrm{FOV}+ N_\textrm{vis} \right),
\end{equation}
and for $w$-projection it is
\begin{equation}
\textrm{TC}_\textrm{wprojection}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} + N_\textrm{vis} w_e^2 \sin^2 \alpha_\textrm{FOV}\right).
\end{equation}
From these bounds it can be concluded that in the limiting behaviour, the $w$-stacking method will be faster when the gridding of the visibilities is the dominating cost of the algorithm. The $w$-projection algorithm will be faster when the inverse FFTs are dominating. In Sect.~\ref{sec:analysis} we will determine which method is faster in practice for different parameters.

@@

\subsection{W-snapshots} \label{sec:snapshot-imaging}
\begin{figure*}
\begin{center}
\includegraphics[width=8cm]{img/vela-normal-projection}
\includegraphics[width=8cm]{img/vela-zenith-projection}
\caption{13-min MWA observation of supernova remnants Vela and Puppis A. Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging. Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm. No beam correction was applied. Computing cost was 60 min for the normal projected image and 41 min for the recentred image.}
\label{fig:vela-projection-example}
\end{center}
\end{figure*}
As discussed, all-sky snapshot imaging with zenith as phase centre is quite efficient. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
%When the Fourier transform is performed, the $w$ term has been corrected (either by w-stacking or by w-projection). Therefore, if $U(u, v)$ is the input to the 2-dimensional Fourier transform, the transform is described by
\begin{align}\notag
\frac{I'(\hat{l}, \hat{m})\left(w_e - w_0\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \times \\
\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging. This essentially is a method of implementing the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}. By doing the inverse corrections during prediction, such a shifted visibility set can be cleaned with Cotton-Schwab iterations similar to a non-shifted set.

>>>>

\def\ANU{$^{1}$}
\def\CAASTRO{$^{2}$}
\def\ASU{$^{3}$}
\def\CASS{$^{4}$}
\def\Curtin{$^{5}$}
\def\CfA{$^{6}$}
\def\Haystack{$^{7}$}
\def\MIT{$^{8}$}
\def\NRAO{$^{9}$}
\def\RRI{$^{10}$}
\def\SKASA{$^{11}$}
\def\Tata{$^{12}$}
\def\UMelbourne{$^{13}$}
\def\USydney{$^{14}$}
\def\UW{$^{15}$}
\def\UWisc{$^{16}$}
\def\Victoria{$^{17}$}

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\hspace{2cm}E-mail:
\url{andre.offringa@anu.edu.au}},
B.~McKinley\ANU$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
F.~H.~Briggs$^{1,2}$, \newauthor
J.~Rhee\ANU,
L.~Feng,
D.~L.~Kaplan\UWisc,
A.~R.~Neben,
R.~B.~Wayth\Curtin$^,$\CAASTRO, \newauthor
J.~D.~Hughes\ANU,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU,
F.~Briggs\ANU,
R.~J.~Cappallo\Haystack,
B.~E.~Corey\Haystack, \newauthor
A.~A.~Deshpande\RRI,
D.~Emrich\Curtin,
B.~M.~Gaensler\USydney$^,$\CAASTRO,
R.~Goeke\MIT,
L.~J.~Greenhill\CfA, \newauthor
B.~J.~Hazelton\UW,
J.~N.~Hewitt\MIT,
M.~Johnston-Hollitt\Victoria,
J.~C.~Kasper\CfA, \newauthor
E.~Kratzenberg\Haystack,
C.~J.~Lonsdale\Haystack,
M.~J.~Lynch\Curtin,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, \newauthor
M.~F.~Morales\UW,
E.~Morgan\MIT,
D.~Oberoi\Tata,
S.~M.~Ord\Curtin$^,$\CAASTRO,
T.~Prabu\RRI, \newauthor
A.~E.~E.~Rogers\Haystack,
D.~A.~Roshi\NRAO \footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
N.~Udaya~Shankar\RRI,
K.~S.~Srivani\RRI,
R.~Subrahmanyan\RRI$^,$\CAASTRO, \newauthor
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
M.~Waterson\Curtin$^,$\ANU,
R.~L.~Webster\UMelbourne$^,$\CAASTRO,
A.~R.~Whitney\Haystack, \newauthor
A.~Williams\Curtin,
C.~L.~Williams\MIT
\\
$^{1}$Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
$^{2}$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
$^{3}$School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
$^{4}$CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
$^{5}$International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
$^{6}$Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
$^{7}$MIT Haystack Observatory, Westford, MA 01886, USA\\
$^{8}$Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
$^{9}$National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
$^{10}$Raman Research Institute, Bangalore 560080, India\\
$^{11}$Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
$^{12}$National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
$^{13}$School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
$^{14}$Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
$^{15}$Department of Physics, University of Washington, Seattle, WA 98195, USA\\
$^{16}$Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
$^{17}$School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
}

@@

\begin{abstract}
We present a new imager implementation that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm. On MWA data, we find it to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution. The performance dependencies of CASA's $w$-projection and WSClean are analysed and analytical functions are derived that describe the required computing cost for both imagers. The difference in computing cost is mainly caused by the efficiency of our new imager on MWA's 27\degree field of view. We predict the computing costs for several other configurations and estimate that our imager is approximately a factor of 2--3 faster on other arrays. Imaging SKA observations is challenging, and SKA1 low will require 3--60 PetaFLOPS of computing power with current techniques. We find that the $w$-snapshot algorithm does not further improve computing requirements. The source code of our new imager is publicly released.
\end{abstract}

@@

Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with a two-dimensional Fast Fourier Transformation (FFT). Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}. This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes that by nature are wide-field instruments.

There are several methods to deal with the $w$-term during imaging: facetting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, $w$-projection \citep{wprojection-cornwell}, $w$-stacking \citep{ska-memo-regridding-2011} and $w$-snapshots \citep{widefield-imaging-ska-cornwell}.

New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}). The Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution. However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's $w$-projection algorithm. Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.

Another option exists for imaging MWA data: the Real-Time System (RTS, \citealt{rts-mwa}). This has been designed as an efficient calibration and imaging pipeline specifically for MWA data. It can use GPUs to improve efficiency, and uses a $w$-projection kernel during imaging to correct for the $w$-term. The kernel can be kept small, because the data is processed in snapshots that have been phase-centered to zenith. The RTS can not perform deconvolution as of yet.

To reach high dynamic ranges, it can be necessary to deal with direction-dependent effects (DDEs). This is especially true for widefield telescopes. One way to handle DDEs is by using the a-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
One such DDE is the effect of the ionosphere. For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure \citep{lonsdale-calibration-approaches-ionosphere}. This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data. The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, $w$-projection and $w$-stacking \citep{awimager-2013}.

Once the Square-Kilometre Array (SKA) begins its operation, the required computational power for wide-field imaging will become an even bigger challenge. \citet{widefield-imaging-ska-cornwell} argues that the $w$-snapshots algorithm is the most efficient approach for the SKA.

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's $w$-projection implementation, and is able to succesfully image any MWA observation. To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with $w$-snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''. Our new imaging implementation, which we believe to be the fastest generic widefield imager available, is publicly released\footnote{TODO add URL}.

@@

\section{The $w$-stacking technique} \label{sec:wstacking}
In this section, we will describe the $w$-stacking algorithm from a mathematical point of view.

@@

To derive the $w$-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \times \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \times \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \times \\ \notag
& e^{2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_{\min}$ to $w_{\max}$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_{\max} - w_{\min}\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \times \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~grid samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~apply the direction dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeat this for all $w$-values and add the results together; v)~apply the final scaling.

In practice, the final scaling will be different from $\left(w_{\max} - w_{\min}\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples. Therefore, the number of added samples and their weight have to be taken into account. Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction dependent effects. In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds. The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-values samples. In this case we can therefore calculate the image for $w<0$ from the image with $w>0$. That allows us to set $w_{\min}$ to the minimum absolute $w$-value, which requires two times fewer layers. In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence the inverse FFT is always performed as a complex-to-complex transform.

@@

\subsection{Discretization of $w$} \label{sec:gridding-w}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/aliasing-example-annotated}
\caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field. WSClean was set to use 12 $w$-layers. The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers. Sources are 1~Jy (red circles), ghost sources are approximately 0.2~Jy. Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce more ghost sources. }
\label{fig:aliasing-example}
\end{center}
\end{figure}
While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of $w$-layers that need to be processed. For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian. This results in the constraint
\begin{equation} \label{eq:minimum-w-distance}
\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
\end{equation}
This suggest that a uniform discretization in $w$ is optimal (TODO this is in contrast to what Cornwell 2008 says -- Cornwell mailed me he made a mistake... Do we need to explicitly point that out here :-/ ?). From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
\begin{equation} \label{eq:nwlayers-bound}
 N_\textrm{wlayers} \gg 2\pi \left(w_{\max} - w_{\min}\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
\end{equation}
Actual values for the right-hand side can differ a lot. The value of $w_{\max} - w_{\min}$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image. For the MWA, a typical value of $w_{\max} - w_{\min}$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$. For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$. This implies that tens of $w$-layers are required at zenith and hundreds at lower elevations. The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in the next sections.

@@

\subsection{Time complexity of $w$-stacking} \label{sec:time-complexity}
It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation. We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each dimension, $N_\textrm{vis}$ is the number of visibilities and $N_\textrm{kern}$ is the size of the antialiasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.

\begin{table}
 \caption{Scaling of the computational cost for various imaging steps} \label{tbl:computational-cost-per-operation}
 \begin{tabular}{lll}
   \textbf{Operation} & \textbf{$w$-stacking} & \textbf{$w$-projection} \\
   \hline\hline
   Fourier transform(s) & $N_\textrm{wlay} N^2_\textrm{pix} \log N_\textrm{pix}$ & $N^2_\textrm{pix} \log N_\textrm{pix}$ \\
   $w$-term corrections   & $N_\textrm{wlay} N^2_\textrm{pix}$ & $N_\textrm{vis} N^2_\textrm{wkern}$ \\
   Gridding & $N_\textrm{vis}N^2_\textrm{kern}$ & $N_\textrm{vis} N_\textrm{kern}^2$ \\
   \hline\hline
 \end{tabular}
\end{table}

Table~\ref{tbl:computational-cost-per-operation} shows how the computational costs scale for the operations that dominate the imaging in the $w$-stacking and $w$-projection algorithms. For comparison, we can assume the gridding kernel adds a constant term and the terms $N_\textrm{wlay}$ and $N_\textrm{wkern}$ follow approximately $w_{\max} \sin \alpha_\textrm{FOV}$. The time complexity for $w$-stacking is then given by
\begin{equation}
\textrm{TC}_\textrm{wstacking}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} w_{\max} \sin \alpha_\textrm{FOV}+ N_\textrm{vis} \right),
\end{equation}
and for $w$-projection it is
\begin{equation}
\textrm{TC}_\textrm{wprojection}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} + N_\textrm{vis} w_{\max}^2 \sin^2 \alpha_\textrm{FOV}\right).
\end{equation}
From these bounds it can be concluded that in the limiting behaviour, the $w$-stacking method will be faster when the gridding of the visibilities is the dominating cost of the algorithm. The $w$-projection algorithm will be faster when the inverse FFTs are dominating. In Sect.~\ref{sec:analysis} we will determine which method is faster in practice for different parameters.

@@

\subsection{$w$-snapshots} \label{sec:snapshot-imaging}
\begin{figure*}
\begin{center}
\includegraphics[width=8cm]{img/vela-normal-projection}
\includegraphics[width=8cm]{img/vela-zenith-projection}
\caption{13-min MWA observation of supernova remnants Vela and Puppis A. Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging. Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm. No beam correction was applied. Computing cost was 60 min for the normal projected image and 41 min for the recentred image.}
\label{fig:vela-projection-example}
\end{center}
\end{figure*}
As discussed, all-sky snapshot imaging with zenith as phase centre is quite efficient. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
%When the Fourier transform is performed, the $w$ term has been corrected (either by $w$-stacking or by $w$-projection). Therefore, if $U(u, v)$ is the input to the 2-dimensional Fourier transform, the transform is described by
\begin{align}\notag
\frac{I'(\hat{l}, \hat{m})\left(w_{\max} - w_{\min}\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \times \\
\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging. This essentially is a method of implementing the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}. By doing the inverse corrections during prediction, such a shifted visibility set can be cleaned with Cotton-Schwab iterations similar to a non-shifted set.
